---
title: "NOAA Storm - Top 10 events to human damage and economic consequences"
author: "Daniel Wong"
date: "January 18, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data Processing
    


```{r}
library(dplyr)
  #read data set - data downloaded from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

    setwd("C://Users//WongD//Documents//R//R-3.3.1//r programming//reproducible_research_p2")
    storm_data<-read.csv("repdata%2Fdata%2FStormData.csv.bz2")
    
    #storm data documentations available at https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
    #and https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf
    
  #data cleansing
    #change all spelling to uppercase for easier grouping
    storm_data$EVTYPE<-toupper(storm_data$EVTYPE)
    
    EVTYPE_table<-table(storm_data$EVTYPE)
    
    #deleting summary lines
    storm_data_revised<-storm_data[grep(".*summary.*", ignore.case=TRUE, storm_data$EVTYPE, invert=TRUE),]
    
    #correct misspellings
    storm_data_revised$EVTYPE<-gsub("AVALANCE","AVALANCHE",storm_data_revised$EVTYPE)
    storm_data_revised$EVTYPE<-gsub("EROSIN","EROSION",storm_data_revised$EVTYPE)
    storm_data_revised$EVTYPE<-gsub("DUST DEVEL","DUST DEVIL",storm_data_revised$EVTYPE)
    storm_data_revised$EVTYPE<-gsub("THUNDERTSORM","THUNDERSTORM",storm_data_revised$EVTYPE)
    storm_data_revised$EVTYPE<-gsub("THUNDESTORM","THUNDERSTORM",storm_data_revised$EVTYPE)
    storm_data_revised$EVTYPE<-gsub("THUNERSTORM","THUNDERSTORM",storm_data_revised$EVTYPE)
    storm_data_revised$EVTYPE<-gsub("TUNDERSTORM","THUNDERSTORM",storm_data_revised$EVTYPE)
    storm_data_revised$EVTYPE<-gsub("TORNDAO","TORNADO",storm_data_revised$EVTYPE)
    storm_data_revised$EVTYPE<-gsub("MICOBURST","MICROBURST",storm_data_revised$EVTYPE)  
    
    #create column event cat based on EVTYPE 
      #create empty column to tag
      event_cat=data.frame(matrix(NA, nrow=nrow(storm_data_revised), ncol=1))
      names(event_cat)<-c("event_cat")
      storm_data_revised<-cbind(storm_data_revised, event_cat)
      
      #read event list created from section 2 of storm documentation.
      #file Event_cat.txt has also been uploaded to github for review.
      event_list<-read.csv("event_cat.txt")
      event_list$event<-toupper(event_list$event)
      
      #tagging by the following logic
      for (j in 1:nrow(event_list)) {
        storm_data_revised$event_cat<-ifelse(grepl(paste("^", event_list$event[j], sep="", collapse="|"),storm_data_revised$EVTYPE),event_list$event[j],storm_data_revised$event_cat)
      }
      
      #keeping only useful fields
      storm_data_revised <- storm_data_revised[,c('EVTYPE','FATALITIES','INJURIES', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP', 'event_cat')]
      
      # Convert H, K, M, B units to calculate Property Damage 
      storm_data_revised$PROPDMGNUM = 0
      storm_data_revised[storm_data_revised$PROPDMGEXP == "H", ]$PROPDMGNUM = storm_data_revised[storm_data_revised$PROPDMGEXP == "H", ]$PROPDMG * 10^2
      storm_data_revised[storm_data_revised$PROPDMGEXP == "K", ]$PROPDMGNUM = storm_data_revised[storm_data_revised$PROPDMGEXP == "K", ]$PROPDMG * 10^3
      storm_data_revised[storm_data_revised$PROPDMGEXP == "M", ]$PROPDMGNUM = storm_data_revised[storm_data_revised$PROPDMGEXP == "M", ]$PROPDMG * 10^6
      storm_data_revised[storm_data_revised$PROPDMGEXP == "B", ]$PROPDMGNUM = storm_data_revised[storm_data_revised$PROPDMGEXP == "B", ]$PROPDMG * 10^9
      
      # Convert H, K, M, B units to calculate Crop Damage
      storm_data_revised$CROPDMGNUM = 0
      storm_data_revised[storm_data_revised$CROPDMGEXP == "H", ]$CROPDMGNUM = storm_data_revised[storm_data_revised$CROPDMGEXP == "H", ]$CROPDMG * 10^2
      storm_data_revised[storm_data_revised$CROPDMGEXP == "K", ]$CROPDMGNUM = storm_data_revised[storm_data_revised$CROPDMGEXP == "K", ]$CROPDMG * 10^3
      storm_data_revised[storm_data_revised$CROPDMGEXP == "M", ]$CROPDMGNUM = storm_data_revised[storm_data_revised$CROPDMGEXP == "M", ]$CROPDMG * 10^6
      storm_data_revised[storm_data_revised$CROPDMGEXP == "B", ]$CROPDMGNUM = storm_data_revised[storm_data_revised$CROPDMGEXP == "B", ]$CROPDMG * 10^9
      
      #total economic consequence = property + crop damage
      storm_data_revised$Total = storm_data_revised$CROPDMGNUM +storm_data_revised$PROPDMGNUM
      
```

Results

```{r}
  #1.Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
  fatal_sum<-aggregate(FATALITIES ~ event_cat, storm_data_revised, FUN = sum, na.rm=FALSE)
  ##head(arrange(fatal_sum,desc(FATALITIES)), n = 10)
  fatal10<-fatal_sum[order(-fatal_sum$FATALITIES),][1:10, ]
  par(las=2)
  par(mar = c(9,8, 1, 1), mgp = c(6, 1.5, 1))
  barplot(fatal10$FATALITIES, main="# of Fatalities by Event (top 10)",horiz= TRUE,names.arg=fatal10$event_cat, xlab="number of fatalities", ylab="event", cex.names=0.6)
  
  injuries_sum<-aggregate(INJURIES ~ event_cat, storm_data_revised, FUN = sum, na.rm=FALSE)
  ##head(arrange(injuries_sum,desc(INJURIES)), n = 10)
  injuries10<-injuries_sum[order(-injuries_sum$INJURIES),][1:10, ]
  par(mar = c(9,8, 1, 1), mgp = c(6, 1.5, 1))
  barplot(injuries10$INJURIES, main="# of  Injuries by Event (top 10)",horiz= TRUE,names.arg=injuries10$event_cat, xlab="number of injuries", ylab="event", cex.names=0.6)
  

  
  
  #2.Across the United States, which types of events have the greatest economic consequences?
  total_cost<-aggregate(Total ~ event_cat, storm_data_revised, FUN = sum, na.rm=FALSE)
  total_cost$Total<-total_cost$Total/1000000
  ##head(arrange(total_cost,desc(Total)), n = 10)
  cost10<-total_cost[order(-total_cost$Total),][1:10,]
  par(mar = c(9,8, 1, 1), mgp = c(6, 1.5, 1))
  barplot(cost10$Total, main="Economic Cost by Event (top 10)",horiz= TRUE,names.arg=cost10$event_cat, xlab="economic cost (in Millions)", ylab="event", cex.names=0.6)
```

Across the United States, events that were most harmful with respect to population health were tornados by far, in terms of both fatalities and injuries. Deaths and injuries attributed to tornados is much higher than other events in the top 10 lists. The event that has the greatest economic consequences are floods, hurricanes, tornados, and other variations of storms, with a noted exception of drought on the list.
